Persistence in functional topology: q-tameness, local-connectivity and a correction to a theorem of Morse

### 2021-04-01 counterexample.tex
Change counterexample: statement using conditions on H-theory

\begin{figure}[t]
	\centering
	\begin{tikzpicture}[scale=88]
	\draw[thick] (-.1,0) rectangle (.1,.05);
	\clip (-.1,0) rectangle (.1,.05); % remove for all circles
	\foreach \i in {1,...,100}{
		\draw[line width=0.1/\i mm] (0, 1/\i^2) circle (1/\i^2);
	}
	\end{tikzpicture}
	\caption{TBW}
	\label{fig:earrings}
\end{figure}

In this section we show that a sublevel filtration can be $\HLC$ and not q-tame.

The $d$-dimensional Hawaiian earring is defined as the subspace
\begin{equation*}
\HE = \bigcup_{n\in\mathbb{N}}\left\{(x_0,\dots,x_d)\in\R^{d+1}\mid \left(x_0-\frac{1}{n}\right)^2+x_1^2+\dots+x_d^2=\left(\frac{1}{n}\right)^2\right\}
\end{equation*}
of $d+1$-dimensional Euclidean space.
Let $f \colon \HE\to\R$ as $0$ at the origin $\vec{0}$ and $1$ everywhere else.
All sublevel sets $\HE_{\leq t}$ are either the singleton $\{\vec{0}\}$ or the whole space $\HE$, and therefore compact.
Moreover, $\HE$ is locally-$f$-connected as we prove now.

Let $e > 0$ and $p \in \HE$.
If $p$ is the origin, we choose $\delta$ between 0 and $\min\{e,1\}$.
Then $\HE_{\leq f(p) + \delta} = \HE_{\leq \delta} = \{p\}$, so the condition is trivially satisfied.
For $p$ not the origin, there is a unique $d$-sphere that contains it.
Clearly, we may choose $\delta$ so small that $B_{\delta}(p)$ contained in this sphere, so $B_{\delta}(p)$ is a disk and the condition follows trivially again.

What remains to be shown is that $\HE_{\leq \bullet}$ is not q-tame with respect to singular and \v{C}ech homology.
Note that $\HE_{\leq t}$ is constant with value $\HE$ for $t \geq 1$, so it suffices to show that the singular and \v{C}ech homology of $\HE$ with field coefficients is infinite dimensional.
The case of singular homology is well know.
For \v{C}ech homology we use the fact that it commutes with inverse limits for compact Hausdorff spaces.
Define 
\begin{align*}
\HE_k &= \left\{(x_0,\dots,x_d)\in\R^{d+1}\mid \left(x_0-\frac{1}{k}\right)^2+x_1^2+\dots+x_d^2=\left(\frac{1}{k}\right)^2\right\}\\
&\cup\bigcup_{n=1}^{k-1}\left\{(x_0,\dots,x_d)\in\R^{d+1}\mid \left(x_0-\frac{1}{n}\right)^2+x_1^2+\dots+x_d^2=\left(\frac{1}{n}\right)^2\right\},
\end{align*}
i.e., the $d$-dimensional Hawaiian earring but with the $k$-th largest $d$-sphere filled.
We have $\lim_{k}\HE_{k} = \bigcap_{k}\mathbb{H}^{d}_{k}=\mathbb{H}^{d}$, and hence $\CH_{d}(\HE) = \lim_{k}\CH_{d}(\mathbb{H}^{d}_{k})$.
One can easily check that each $\mathbb{H}^{d}_{k}$ satisfies the assumptions for \cref{prop:cech_sing_hom_hlc}, which implies $\lim_{k}\CH_{d}(\mathbb{H}^{d}_{k})=\lim_{k}H_{d}(\mathbb{H}^{d}_{k})$.
We compute
\[
\lim_{k}H_{d}(\mathbb{H}^{d}_{k})=\lim\left(\dots\to \prod_{n=1}^2\mathbb{F}\to \prod_{n=1}^1\mathbb{F}\to \prod_{n=1}^0\mathbb{F}\right)=\prod_{n\in\mathbb{N}}\mathbb{F},
\]
which is infinite-dimensional over $\mathbb{F}$.
This finishes the proof.

\begin{rem}
Note that the singular persistent homology of the sublevel set filtration of $F$ above is also not q-tame because the $d$-dimensional Hawaiian earring has infinite-dimensional $d$-th singular homology \cite{Barratt.1962}. 
\end{rem}

\begin{rem}
The construction above does not yield a counterexample to \cref{con:q-tame_pers_cech_hom} because $\mathbb{H}^{d}$ is not strongly locally-$F$-connected at the origin.
\end{rem}

>>>>

\begin{figure}[t]
	\centering
	\begin{tikzpicture}[scale=88]
	\draw[thick] (-.1,0) rectangle (.1,.05);
	\clip (-.1,0) rectangle (.1,.05); % remove for all circles
	\foreach \i in {1,...,100}{
		\draw[line width=0.1/\i mm] (0, 1/\i^2) circle (1/\i^2);
	}
	\end{tikzpicture}
	\caption{A closeup of the Hawaiian hearing $\mathbb{H}^1$.}
	\label{fig:earrings}
\end{figure}

In this section we show that a sublevel filtration can be $\HLC$ and not q-tame for most homology theories of interest.

The \textit{$d$-dimensional Hawaiian earring}, see Figure \ref{fig:earrings}, is defined as the subspace
\begin{equation*}
\HE = \bigcup_{n\in\mathbb{N}}\left\{(x_0,\dots,x_d)\in\R^{d+1}\mid \left(x_0-\frac{1}{n}\right)^2+x_1^2+\dots+x_d^2=\left(\frac{1}{n}\right)^2\right\}
\end{equation*}
of $d+1$-dimensional Euclidean space.

\begin{thm} \label{thm:counterexample}
	Let $\H$ be a homology theory such that its value on a point is finitely generated for every degree and on $\HE$ is not finitely generated for some degree. Then, the function $f \colon \HE \to \R$ whose value at the origin is $0$ and is $1$ everywhere else defines a $\HLC$ Morse filtration that is not q-tame.
\end{thm}

\begin{proof}
	The space $\HE$ is Hausdorff and locally compact since $R^{d+1}$ is.
	To verify that $(\HE,f)$ is a Morse filtration we notice that all level sets are either the empty set, the singleton containing the origin, or $\HE$ itself, all compact spaces.
	Let us now verify that $\HE$ is $\HLC$.
	Let $\epsilon > 0$ and $x \in \HE$.
	If $x$ is the origin, we choose $\delta$ between $0$ and the minimum of $\epsilon$ and $1$.
	Then $\HE_{\leq f(x) + \delta} = \HE_{\leq \delta} = \{x\}$, so the desired condition follows from the first assumption on the homology theory.
	For $x$ not the origin, there is a unique $d$-sphere that contains it.
	Clearly, we may choose $\delta > 0$ so small that $D_\delta(x) = \{y \in \R^{d+1} \mid \Vert x - y \Vert < \delta\} \cap \HE$ is contained in this sphere, so $D_\delta(x)$ is homotopy equivalent to a $x$ and the condition again follows trivially.
	What remains to be shown is that $\HE_{\leq \bullet}$ is not q-tame.
	Since $\HE_{\leq t}$ is constant with value $\HE$ for $t \geq 1$, the second assumption on the homology theory finishes the proof.
\end{proof}

\begin{prop}
	The assumptions of Theorem~\ref{thm:counterexample} are satisfied by singular and \v{Cech} homology.
\end{prop}

\begin{proof}
	The case of singular homology is well know and can be found in \cite{Barratt.1962}.
	For \v{C}ech homology we use the fact that it commutes with inverse limits for compact Hausdorff spaces.
	Define 
	\begin{align*}
	\HE_k &= \left\{(x_0,\dots,x_d)\in\R^{d+1}\mid \left(x_0-\frac{1}{k}\right)^2+x_1^2+\dots+x_d^2=\left(\frac{1}{k}\right)^2\right\}\\
	&\cup\bigcup_{n=1}^{k-1}\left\{(x_0,\dots,x_d)\in\R^{d+1}\mid \left(x_0-\frac{1}{n}\right)^2+x_1^2+\dots+x_d^2=\left(\frac{1}{n}\right)^2\right\},
	\end{align*}
	i.e., the $d$-dimensional Hawaiian earring but with the $k$-th largest $d$-sphere filled.
	We have $\lim_{k}\HE_{k} = \bigcap_{k}\mathbb{H}^{d}_{k}=\mathbb{H}^{d}$, and hence $\CH_{d}(\HE) = \lim_{k}\CH_{d}(\mathbb{H}^{d}_{k})$.
	One can easily check that each $\mathbb{H}^{d}_{k}$ satisfies the assumptions for \cref{prop:cech_sing_hom_hlc}, which implies $\lim_{k}\CH_{d}(\mathbb{H}^{d}_{k})=\lim_{k}H_{d}(\mathbb{H}^{d}_{k})$.
	We compute
	\begin{equation*}
	\lim_{k}H_{d}(\mathbb{H}^{d}_{k})=\lim\left(\dots\to \prod_{n=1}^2\mathbb{F}\to \prod_{n=1}^1\mathbb{F}\to \prod_{n=1}^0\mathbb{F}\right)=\prod_{n\in\mathbb{N}}\mathbb{F},
	\end{equation*}
	which is infinite-dimensional over $\mathbb{F}$.
	This finishes the proof.
\end{proof}

### 2021-04-01 introduction.tex
Split preliminaries
Change counterexample: statement using conditions on H-theory

\begin{defi}
	Let $M$ be a metric space and $f \colon M \to \R$ a function. The space $M$ is called \emph{locally-$f$-connected} if for each $p\in M$, $e>0$ and $d\in\mathbb{N}_0$ there exists $\delta\in(0,e)$ such that the map 
	\[
	\tilde{H}_d(f_{\leq f(p)+\delta}\cap B_{\delta}(p)\hookrightarrow f_{\leq f(p)+e}\cap B_e(p);\mathbb{Z})
	\]
	is 0.
\end{defi}

@@

>>>>

\anibal{I would quote the original verbatim}
\begin{defi}
	Let $M$ be a metric space and $f \colon M \to \R$ a function. The space $M$ is called \emph{locally-$f$-connected} if for each $p\in M$, $e>0$ and $d\in\mathbb{N}_0$ there exists $\delta\in(0,e)$ such that the map 
	\[
	\tilde{H}_d(f_{\leq f(p)+\delta}\cap B_{\delta}(p)\hookrightarrow f_{\leq f(p)+e}\cap B_e(p);\mathbb{Z})
	\]
	is 0.
\end{defi}

@@

\anibal{recall that this is not equivalent to Morse's condition.
	I would not write down the condition below but refer to the definition in the theorem section.
	Here I think is better to explain in words sating that we were inspired a condition stronger than the one above that is stable under forward translation.}

### 2021-04-01 theorem.tex
Change counterexample: statement using conditions on H-theory

In this section we introduce a stronger version of $\HLC$ for sublevel filtrations, and we show that Morse filtrations with this property are q-tame.

@@

\begin{thm} \label{t:strong local connectenss implies q-tameness}
	Any be a strongly $\HLC$ Morse filtration $(X,f)$ is q-tame, i.e., for every $s < t$ the inclusion $X_s \to X_t$ is $\HS$.
\end{thm}

\begin{lem} \label{l:commutative algebra}
	Given a commutative diagram of modules over a principal ideal domain
	\begin{equation*}
	\begin{tikzcd}
	A_{1,1} \arrow[r] & A_{1,2} & \\
	A_{2,1} \arrow[r] \arrow[u] & A_{2,2} \arrow[r] \arrow[u] & A_{2,3} \\
	& A_{3,2} \arrow[r] \arrow[u] & A_{3,3} \arrow[u]
	\end{tikzcd}
	\end{equation*}
	where the middle row is exact and both $A_{2,1} \to A_{1,1}$ and $A_{3,3} \to A_{2,3}$ have finitely generated images, then $A_{3,2} \to A_{1,2}$ does as well.
\end{lem}

\begin{proof}
	This is proven via a straightforward diagram chase. For complete details see Lemma 17.3 in \cite{Bredon.1968}.
\end{proof}

@@

\begin{proof}
	The lemma holds for $\HS$ replaced by $\HS_{(n-1)}$ for any $n \leq 0$ since $\H_{n-1}$ induces the zero map. We will proceed by induction on $n$ assuming the lemma for $\HS_{(n-1)}$.

>>>>

In this section we introduce a stronger version of $\HLC$ for sublevel filtrations and show that Morse filtrations with this property are q-tame.

@@

\begin{thm} \label{t:strong local connectenss implies q-tameness}
	Strongly $\HLC$ Morse filtrations are q-tame for any homology theory.
\end{thm}

\begin{lem} \label{l:commutative algebra}
	Given a commutative diagram of modules over a principal ideal domain
	\begin{equation*}
	\begin{tikzcd}
	A_{1,1} \arrow[r] & A_{1,2} & \\
	A_{2,1} \arrow[r] \arrow[u] & A_{2,2} \arrow[r] \arrow[u] & A_{2,3} \\
	& A_{3,2} \arrow[r] \arrow[u] & A_{3,3} \arrow[u]
	\end{tikzcd}
	\end{equation*}
	where the middle row is exact and both $A_{2,1} \to A_{1,1}$ and $A_{3,3} \to A_{2,3}$ have finitely generated images, then so does $A_{3,2} \to A_{1,2}$.
\end{lem}

\begin{proof}
	This is proven via a straightforward diagram chase. For more details see Lemma 17.3 in \cite{Bredon.1968}.
\end{proof}

@@

\begin{proof}
	We need to prove that the inclusion $X_s \to X_t$ is $\HS$ for every $s < t$.

	The lemma holds for $\HS$ replaced by $\HS_{(n-1)}$ for any $n \leq 0$ since $\H_{n-1}$ induces the zero map. We will proceed by induction on $n$ assuming the lemma for $\HS_{(n-1)}$.
